### 2015-08-01 Harris_2015/manuscript-materials/draft.Rmd
things

*Running head:* Species interactions in Markov networks

**Title: Estimating species interactions from observational data with Markov
networks**

@@

\setlength{\parindent}{1cm}
\setlength{\parskip}{4pt}

@@

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumptions that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@diamond_island_1975; @strong_ecological_1984;
@gotelli_empirical_2009].

@@

Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
presence), a Markov network defines the relative probability of observing a
given presence-absence vector, $\vec{y}$, as

\centering

$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$

\raggedright
\setlength{\parindent}{1cm}

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
(i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the
*rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to define the objective function and gradient in R code
and then find the optimum.

@@

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method) to calibrate the performance of the Markov
network as a method for estimating species interactions. The first two
alternative interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
are undefined for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in the
corpcor package’s `pcor.shrink` function with the default settings
[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products [@REF].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. [[Observational
presence-absence data does not have enough degrees of freedom to estimate this
many parameters, so I symmetrized the pairwise correlations by averaging the
coefficient for predicting species $i$’s status from species $j$ and vice
versa.]]

The fourth alternative method, described in @gotelli_empirical_2009, involved
simulating possible landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. Using the default options
in the Pairs software described in @gotelli_empirical_2009, I simulated the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species). The software then calculated a
$Z$ statistic for each species pair using this null distribution, which I used
as its estimator of species interactions.

@@

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics (Appendix), I also reported
how often they rejected the null hypothesis of no interaction.

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The methods for evaluating conditional
relationships (Markov networks, GLMs, and both types of partial covariance
estimates), however, successfully controlled for the indirect pathway via
the tree species and each correctly identified the direct negative interaction
between the shrubs 94% of the time.

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3a). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (34.6% overall), followed
by the generalized linear models (29.9%), partial correlations from the raw
presence-absence data (27.5%), and partial correlations from BayesComm, the
joint species distribution model (26.1%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [[ref]].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (18.6%), followed by the null model
(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
methods had different $R^2$ values, there was a close mapping between
their estimates (after controlling for the size of the simulated landscapes;
Figure 3b).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the information that could be obtained more easily by
calculating correlation coefficients between species' presence-absence vectors.

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions (e.g. which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can
reciprocally influence one another in multiple ways simultaneously
[@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists would
need other kinds of data, as from time series, behavioral observations,
manipulative experiments, or natural history. These other sources of information
could also be used to augment the likelihood function with an informative prior
distribution, which could lead to better results on some real data sets than
was shown in Figure 3a.

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. In complex ecological networks (and even in
small networks with three species), it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3b).
Estimating pairwise coefficients directly seems like a much more promising
approach: to the extent that the models' relative performance on real data sets
is similar to the range of results shown in Figure 3a, scientists in this field
could often double their explanatory power by switching from null models to
Markov networks (or increase it by nearly as much with linear or generalized
linear approximations). Regardless of the specific methods ecologists ultimately
choose, the most important consideration is clearly that we need to control for
indirect effects and estimate conditional relationships between species, rather
than marginal ones.

@@

>>>>

\chead{\textit{Species interactions in Markov networks}}

**Title:** Estimating species interactions from observational data with Markov
networks

@@

\setlength{\parindent}{1em}
\setlength{\parskip}{3pt}

@@

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumptions that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

@@

Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
presence), a Markov network defines the relative probability of observing a
given presence-absence vector, $\vec{y}$, as
$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
(i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the
*rosalia* package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters.

@@

I also evaluated five alternative methods from the existing literature (plus a
novel variant of the fifth method) to calibrate the performance of the Markov
network as a method for estimating species interactions. The first two
alternative interaction measures were the sample correlations and the partial
correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
are undefined for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in the
corpcor package’s `pcor.shrink` function with the default settings
[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. To avoid the
identifiability problems associated with directed cyclic graphs
[@schmidt_modeling_2012], I then symmetrized the pairwise relationships within
species pairs via averaging.

The fourth alternative method, described in @gotelli_empirical_2009, involved
simulating possible landscapes from a null model that retains the row and column
sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
computed by the Pairs software described in @gotelli_empirical_2009 as my
null model-based estimator of species interactions.

@@

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed (Appendix). For the null
model (Pairs), I calculated statistical significance using its $Z$-score. For
the Markov network, I used the Hessian matrix to generate approximate confidence
intervals and noted whether these intervals included zero.

@@

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the correlation
between these species was positive in 94% of replicates, and the
randomization-based null model falsely reported positive associations 100% of
the time. Worse, more than 98% of these false conclusions were statistically
significant. The partial correlation and Markov network estimates, on the other
hand, each correctly identified the direct negative interaction between the
shrubs 94% of the time (although the confidence intervals overlapped zero in
most of the replicates).

#### Twenty species.
The accuracy of the evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3A). In general,
however, there was a consistent ordering, with the four methods that controlled
for indirect effects of species interactions performing the best: the Markov
network explained the largest portion of the variance (34.6% overall), followed
by the generalized linear models (29.9%), partial correlations from the raw
presence-absence data (27.5%), and partial correlations from BayesComm, the
joint species distribution model (26.1%).  The benefit of choosing the full
Markov network over the other three methods was largest on the smaller
landscapes, which are also the ones that are most representative of typical
analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (18.6%), followed by the null model
(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
methods had different $R^2$ values, there was a close mapping between
their estimates (after controlling for the size of the simulated landscapes;
Figure 3B).  This suggests that the correlation matrices from joint species
distribution models and especially the effect sizes from null models only
contain noisy versions of the information that could be obtained more easily by
calculating correlation coefficients between species' presence-absence vectors.

@@

It’s important to note that none of these methods can identify the exact nature
of the pairwise interactions (e.g. which species in a positively-associated
pair is facilitating the other), particularly when real pairs of species can
reciprocally influence one another in multiple ways simultaneously
[@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists would
need other kinds of data, as from time series, behavioral observations,
manipulative experiments, or natural history. These other sources of information
could also be used to augment the likelihood function with an informative prior
distribution, which could lead to better results on some real data sets than
was shown in Figure 3A.

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
particular null model must be interpreted with care
[@roughgarden_competition_1983]. In complex ecological networks (and even in
small networks with three species), it may simply not be possible to implicate
individual species pairs or specific ecological processes like competition by
rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
test statistic is effectively just a correlation coefficient (Figure 3B).
Estimating pairwise coefficients directly seems like a much more promising
approach: to the extent that the models' relative performance on real data sets
is similar to the range of results shown in Figure 3A, scientists in this field
could often double their explanatory power by switching from null models to
Markov networks (or increase it by nearly as much with linear or generalized
linear approximations). Regardless of the specific methods ecologists ultimately
choose, the most important consideration is clearly that we need to control for
indirect effects and estimate conditional relationships between species, rather
than marginal ones.

@@

\setlength{\parskip}{2pt}

### 2015-08-01 Harris_2015/simulation.Rmd
things

```{r}
library(progress)
set.seed(1)
```

@@

$$ $$

@@

```{r}

simulate_communities = function(n_spp = 20,
         K = n_spp / 3,
         maxit = 5E6,
         thin = 1E4,
         seed_rain = rep(.1, n_spp),
         growth = rep(1, n_spp),
         interaction_strength = 1/2,
         prob_negative = 0.75,
         mutualism_maximum = 0.5
){

@@

```{r}
thin = 5E3
n_spp = 20
n_reps = 50

for(n_sites in c(25, 200, 1600)){
  message(n_sites)

>>>>

```{r}
library(progress)
set.seed(1)

thin = 5E3
n_spp = 20
n_reps = 50
```

@@

(1 + exp(0.2 * x / K / n_spp)) / 2

$$f(\vec{n} = $$

@@

```{r}
simulate_communities = function(n_spp = 20,
         K = n_spp / 3,
         maxit = 5E6,
         thin = 1E4,
         seed_rain = rep(.1, n_spp),
         growth = rep(1, n_spp),
         interaction_strength = 1/2,
         prob_negative = 0.75,
         mutualism_maximum = 0.5
){

@@

```{r, eval=FALSE}
for(n_sites in c(25, 200, 1600)){
  message(n_sites)

### 2015-08-01 Harris_2015/three_species.Rmd
ready to submit, maybe?
things

```{r}
library(dplyr)
library(magrittr)
library(rosalia)
library(knitr)
library(rosalia)
```

@@

Once a landscape has been simulated, test whether `rosalia`, the sample covariance, and the partial 
covariance got the sign of the shrub-shrub interaction correct.
```{r}
evaluation = sapply(
  1:1000,
  function(i){
    x = rosalia:::simulate_data(
      par = truth, 
      possibilities = possibilities, 
      possible_cooc = possible_cooc, 
      n_sites = n_sites
    )
    if(paste0("fakedata/three/", i, ".txt") %in% dir("fakedata/three", full.names = TRUE)){
      # Do nothing: simulated data was already written to disk
    }else{
      write.table(t(x[rowSums(x) > 0, ]), file = paste0("fakedata/three/", i, ".txt"))
    }

    c(
      cov = cov(x)[2, 3] < 0,
      partial = -solve(cov(x))[2, 3] < 0,
      rosalia = rosalia(x, trace = 0, hessian = FALSE)$par[[3]] < 0
    )
  })

@@

# Find the right rows of the text file, split them on whitespace, pull out
# the column identified above, and find its sign
pairs_evaluation = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
    strsplit(" +") %>%
    sapply(. %>% extract(z_score_column) %>% as.numeric) %>%
    is_greater_than(0) # Z>0 implies negative interactions because the Z-score
                       # relates to the distribution of C-scores, which describe
                       # the frequency with which species *do not* co-occur
```

Percentage of correct sign estimates:
```{r, eval=FALSE}
mean(pairs_evaluation) * 100

rowMeans(evaluation) * 100
```

```{r, echo = FALSE}
out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation) * 100)

@@

>>>>

```{r, message = FALSE}
library(dplyr)
library(magrittr)
library(rosalia)
library(knitr)
library(corpcor)
```

@@

Once a landscape has been simulated, test whether `rosalia`, the sample correlation, and the partial 
correlation got the sign of the shrub-shrub interaction correct.

```{r, message=FALSE, cache = TRUE}
evaluation = sapply(
  1:1000,
  function(i){
    x = rosalia:::simulate_data(
      par = truth, 
      possibilities = possibilities, 
      possible_cooc = possible_cooc, 
      n_sites = n_sites
    )
    if(paste0("fakedata/three/", i, ".txt") %in% dir("fakedata/three", full.names = TRUE)){
      # Do nothing: simulated data was already written to disk
    }else{
      write.table(t(x[rowSums(x) > 0, ]), file = paste0("fakedata/three/", i, ".txt"))
    }

    rosie = rosalia(x, trace = 0, prior = make_flat_prior(), hessian = TRUE)

    se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
    rosalia_confident_positive = pnorm(0, mean = rosie$beta[2, 3], sd = se) < .025
    rosalia_confident_negative = pnorm(0, mean = rosie$beta[2, 3], sd = se) > .975

    c(
      cor = cor(x)[2, 3] < 0,
      partial = cor2pcor(cor(x))[2, 3] < 0,
      rosalia = rosie$beta[2,3] < 0,
      rosalia_confident_positive = rosalia_confident_positive,
      rosalia_confident_negative = rosalia_confident_negative
    )
  })

@@

# Find the right rows of the text file, split them on whitespace, pull out
# the column identified above, and save the Z-score
pairs_z = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
    strsplit(" +") %>%
    sapply(. %>% extract(z_score_column) %>% as.numeric)

# Z>0 implies negative interactions because the Z-score relates to the
# distribution of C-scores, which describe the frequency with which species *do
# not* co-occur. It has the opposite sign of metrics based on when species *do*
# co-occur.
pairs_evaluation = pairs_z > 0
```

```{r}
out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation)[1:3] * 100)

@@

```{r}
# percentage of Pairs runs that found statistically significant mutualism at the
# p<.05 level (two-tailed test)
mean(pnorm(pairs_z) < 0.025) * 100

# percentage of Pairs runs that found statistically significant competition at the
# p<.05 level (two-tailed test)
pairs_significant_competition = pnorm(pairs_z) > 0.975

# percentage of rosalia runs whose 95% confidence intervals did not include 0
rowMeans(evaluation)[c("rosalia_confident_negative", "rosalia_confident_positive")] * 100
```
